Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2018-07-15 paper.tex
...

Our case study shows gains in productivity for four selected behaviors in the
game logic of Pingus rewritten in \CEU.
We present an in-depth qualitative analysis of the proposed solutions in
comparison to the original implementations in C++.
%
Not all techniques result in reduction of \locs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
%
We also identify four control-flow patterns that likely apply to most games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
%
In this context, a control-flow pattern is a recurring technique to describe
execution dependency and/or explicit ordering between statements.
% (or groups of statements).
%For instance, consider how a key press stimulus propagates through the game
%entities and also what happens with them if the stimulus causes the end of the
%game.
%We focus on a qualitative analysis for the programming techniques that we
%applied during the rewriting process.

We employed a \emph{live code rewrite}, i.e., starting from the original
codebase in C++, we reimplemented it piece-by-piece in \CEU without breaking
the game compilation and execution.
%
This approach shows the feasibility of a partial and gradual translation
between the languages, and also the possibility of keeping performance-critical
functionality in C++, such as in heavy graphical primitives.

@@

\begin{lstlisting}[
  numbers=right,xrightmargin=0.5cm,
  float=t,
  caption={An example in \CEU that blinks an LED every 500ms in the background
          and awaits a button click or 1 hour to terminate.},
  label={lst.ceu},
]
input  void BUTTON;
output bool LED;

@@

\CEU is a synchronous reactive language in which programs evolve in a sequence
of discrete reactions to external events with the characteristics that follow:
%
\begin{itemize}
\item \textbf{Reactive:} code only executes in reactions to events.
\item \textbf{Structured:} programs use structured control mechanisms, such as
    \code{spawn} and \code{await} (to create and suspend activities).
\item \textbf{Synchronous:} reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{itemize}
%
\CEU is designed for control-intensive applications, supporting concurrent
lines of execution, known as \emph{trails}, and instantaneous broadcast
communication through events.
%
Computations within a reaction (such as expressions, assignments, and system
calls) are also instantaneous considering the synchronous
hypothesis~\cite{rp.hypothesis}.
%
%\CEU provides an \code{await} statement which is the only that actually
%``consumes'' time.
%
\CEU provides an \code{await} statement that blocks the current running trail
allowing the program to execute its other trails; when all trails are blocked,
the reaction terminates and control returns to the environment to process
upcoming events (e.g., mouse clicks and timers).

Listing~\ref{lst.syntax} shows a compact reference of \CEU and
Listing~\ref{lst.ceu} shows a simple example that blinks an LED and terminates
either on a button press or after 1 hour of execution.
%
The example declares two external events (ln~1--2): an input that represents
a button and an output that represents an LED.
%
The blink behavior uses a \code{code} abstraction (ln~4--11) which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}, but which can also
interact directly with the environment through \code{await} statements
(ln.~7,9).
%
The \code{spawn} statement (ln~13) instantiates a \code{Blink} abstraction to
execute in the background and the program then creates two trails to await
two events at the same time (ln~14--18).
Since the composition is a \code{par/or}, it will terminate when either of the
two events occur proceeding to the end of the file and consequently also
terminating the program.

@@

In Pingus, the game logic accounts for almost half the size of the codebase%
\footnote{
We used \emph{SLOCCount} to count only non-blank, non-comment lines in the
codebase: \url{www.dwheeler.com/sloccount/}
}:
18.173 from 39.362 \locs (46\%) spread across 272 files.
%
However, about half of the game logic relates to non-reactive code, such as
dealing with configurations and options, saved games and serialization, maps
and level descriptions, string formatting, collision detection, graph
algorithms, etc.
This part remains unchanged and relies on the seamless integration between \CEU
and C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the
integration happens at the source code level.
This enables accessing data and calling C/C++ from \CEU and vice-versa.
%
Therefore, we only rewrote 9.186 \locs spread across 126 files%
\footnote{\label{codebase} Effective codebase: \url{github.com/an000/p/tree/master/}}.
%
In order to only consider relevant code in the analysis, we then removed all
headers, declarations, trivial getters \& setters, and other innocuous
statements, resulting 4.135 condensed \locs spread across 70 implementation
files originally written in C++%
\footnotemark[\ref{codebase}].
%\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
We did the same with the implementation in \CEU, resulting in 3.697 condensed
\locs%
\footnotemark[\ref{codebase}].
%\footnote{\CEU codebase: \url{github.com/an000/p/tree/master/all}}.
%
Figure~\ref{tab.tree} summarizes the effective game logic codebase in the two
implementations.

@@

The rewriting process consisted of identifying sets of callbacks in C++
implementing control flow in the game and translating them to \CEU using
appropriate structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
As a general rewriting rule, we identify control-flow behaviors in the C++
codebase by looking for class state members with identifiers resembling verbs,
statuses, and counters (e.g.,
\code{pressed},
\code{particle\_thrown},
\code{mode}, and
\code{delay\_count}).
Good chances are that such variables encode some form of control-flow
progression that cross multiple callback invocations.
%
Not all state follows these conventions, but they helped finding classes that
are heavy on control flow quickly at the beginning of the process.

@@

\subsubsection{ Transition from \emph{Story}  to \emph{Credits} and \emph{Worldmap} Screen}

@@

Continuation passing typically controls the overall structure of games in, such
as screen transitions in menus and level progressions.
%
In Pingus, \CEU adopts the direct style technique in five cases involving
screen transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches between pingu
actions during gameplay (e.g., \emph{walking} to \emph{falling} and back to
\emph{walking}).

@@

In C++, the class \code{Bomber} in Listing~\ref{lst.hier.a} declares a
\code{sprite} member (ln 3) to handle its animation frames.
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus to the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{ScreenManager::display} in the main game loop calls\\
      \code{ScreenManager::update} when starting a new frame.
\item \code{ScreenManager::update} calls \code{screen->update} for the active
      game screen (i.e., a \code{GameSession} instance, considering the screen
      in which the \code{Bomber} appears).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{objs->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is indeed necessary considering the typical
OO game architecture:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item With a single assignment to \code{screen}, one can easily deactivate the
current screen and redirect all dispatches to a new screen (step 2).
\item The \code{World} class manages and dispatches events to all game
      entities with a common interface \code{WorldObj}, such as the pingus and
      traps (step 4).
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus (step 5).
\item Since a single pingu can change its actions during lifetime, the
      \code{action} member decouples them with another level of indirection
      (step 6).
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions (step 7).
\end{itemize}
%
Like \code{update}, the \code{draw} callback also flows through a similar
dispatching hierarchy until reaching the \code{Sprite} class.

@@

In \CEU, the \code{Bomber} abstraction presented in Listing~\ref{lst.hier.b}
spawns a \code{Sprite} animation instance on its body (ln 3).
%
However, in \CEU, the \code{Sprite} abstraction can react directly to external
\code{update} and \code{draw} events, bypassing the program hierarchy entirely.
Events in \CEU are broadcasted to the entire application in lexical order,
i.e., an abstraction that appears first in the source code (e.g., ln 3) reacts
before another one that appears second (e.g., hidden in ln 4).
%As discussed in Section~\ref{sec.pats.fsms.2}
This rule preserves determinism and also conforms to the program static
hierarchy.
While (\emph{and only while}) the bomber abstraction is alive, the sprite
animation remains alive and reacts to the \code{update} and \code{draw} events.
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.
% TODO: hidden in the language runtime
%Note that one can still declare a local event to restrict its visibility like
%a local variable.

@@

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
In addition, the \code{spawn} statement supports a pool identifier to associate
a new instance with a pool.
%
The game screen in Listing~\ref{lst.pingus.b} spawns a new \code{Pingu} on every
invocation of \code{Pingu\_Spawn} (ln 4--7).
%
The \code{spawn} statement (ln 6) specifies the pool declared at the top-level
block of the game screen (ln 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln 4--7), surviving the call to \code{Pingu\_Spawn}.
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its \emph{natural termination}, the instance is automatically
removed from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln 17) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

>>>>

Our case study discusses the gains in productivity for four selected behaviors
in the game logic of Pingus rewritten in \CEU.
We present an in-depth qualitative analysis of the proposed solutions in
comparison to the original implementations in C++.
%
Not all techniques result in reduction of \locs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
%
We also identify four control-flow patterns that likely apply to most games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
%
In this context, a control-flow pattern is a recurring technique to describe
execution dependency and/or explicit ordering between statements.
% (or groups of statements).
%For instance, consider how a key press stimulus propagates through the game
%entities and also what happens with them if the stimulus causes the end of the
%game.
%We focus on a qualitative analysis for the programming techniques that we
%applied during the rewriting process.

We employed a \emph{live code rewrite}, i.e., starting from the original
codebase in C++, we reimplemented it piece-by-piece in \CEU without breaking
the game compilation and execution.
%
This approach shows the feasibility of a partial and gradual translation
between the languages, and also the possibility of keeping performance-critical
functionality in C++, such as for heavy graphical primitives.

@@

\begin{lstlisting}[
  numbers=right,xrightmargin=0.5cm,
  float=t,
  caption={An example in \CEU that blinks an LED every 500ms in the background
          and awaits either a button click or 1 hour to terminate.},
  label={lst.ceu},
]
input  void BUTTON;
output bool LED;

@@

\CEU is a synchronous reactive language in which programs evolve in a sequence
of discrete reactions to external events with the characteristics that follow:
%
\begin{itemize}
\item \textbf{Reactive:} code only executes in reactions to events.
\item \textbf{Structured:} programs use structured control mechanisms, such as
    \code{spawn} and \code{await} (to create and suspend activities).
\item \textbf{Synchronous:} reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{itemize}
%
\CEU is designed for control-intensive applications, supporting concurrent
lines of execution, known as \emph{trails}, and instantaneous broadcast
communication through events.
%
%Computations within a reaction (such as expressions, assignments, and system
%calls) are also instantaneous considering the synchronous
%hypothesis~\cite{rp.hypothesis}.
%
%\CEU provides an \code{await} statement which is the only that actually
%``consumes'' time.
%
\CEU provides an \code{await} statement that blocks the current running trail
allowing the program to execute its other trails; when all trails are blocked,
the reaction terminates and control returns to the environment to process
upcoming events.
%
Listing~\ref{lst.syntax} shows a compact reference of \CEU.

Listing~\ref{lst.ceu} shows a simple example in the context of embedded systems
that blinks an LED and terminates either on a button press or after 1 hour of
execution.
%
The example declares two external events (ln 1--2): an input that represents
a button and an output that represents an LED.
%
The blink behavior uses a \code{code} abstraction (ln 4--11) which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}, but which can interact
directly with the environment through \code{await} statements (ln 7,9).
%
The \code{spawn} statement (ln 13) instantiates a \code{Blink} abstraction to
execute in the background and the program then creates two trails to await
two events at the same time (ln 14--18).
Since the composition is a \code{par/or}, it will terminate when either of the
two events occur, proceeding to the end of the file and consequently also
terminating the program.

@@

In Pingus, the game logic accounts for almost half the size of the codebase%
\footnote{
We used \emph{SLOCCount} to count only non-blank, non-comment lines in the
codebase: \url{www.dwheeler.com/sloccount/}
}:
18.173 from 39.362 \locs (46\%) spread across 272 files.
%
However, about half of the game logic relates to non-reactive code, such as
dealing with configurations and options, saved games and serialization, maps
and level descriptions, string formatting, collision detection, graph
algorithms, etc.
This part remained unchanged in our rewrite and relies on the seamless
integration between \CEU and C/C++~\cite{ceu.sensys13}: the type systems are
compatible and the integration happens at the source code level.
This enables accessing data and calling C/C++ from \CEU and vice-versa.
%
Therefore, we only rewrote 9.186 \locs spread across 126 files%
\footnote{\label{codebase} Effective codebase: \url{github.com/an000/p/tree/master/}}.
%
In order to only consider relevant code in the analysis, we then removed all
headers, declarations, trivial getters \& setters, and other innocuous
statements, resulting 4.135 condensed \locs spread across 70 implementation
files originally written in C++%
\footnotemark[\ref{codebase}].
%\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
We did the same with the implementation in \CEU, resulting in 3.697 condensed
\locs%
\footnotemark[\ref{codebase}].
%\footnote{\CEU codebase: \url{github.com/an000/p/tree/master/all}}.
%
Figure~\ref{tab.tree} summarizes the effective game logic codebase in the two
implementations.

@@

The rewriting process consisted of identifying sets of callbacks in C++
implementing control flow in the game and translating them to \CEU using
appropriate structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
As a general rewriting rule, we identify control-flow behaviors in the C++
codebase by looking for class state members with identifiers resembling verbs,
statuses, and counters (e.g.,
\code{pressed},
\code{particle\_thrown},
\code{mode}, and
\code{delay\_count}).
Good chances are that such variables encode some form of control-flow
progression that crosses multiple callback invocations.
%
Not all state follows these conventions, but they helped finding classes that
are heavy on control flow quickly at the beginning of the process.

@@

\subsubsection{ Transition from \emph{Story}  to \emph{Credits} to \emph{Worldmap} Screen}

@@

Continuation passing typically controls the overall structure of games such as
screen transitions in menus and level progressions.
%
In Pingus, \CEU adopts the direct style technique in five cases involving
screen transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches between pingu
actions during gameplay (e.g., \emph{walking} to \emph{falling} and back to
\emph{walking}).

@@

In C++, the class \code{Bomber} in Listing~\ref{lst.hier.a} declares a
\code{sprite} member (ln 3) to handle its animation frames.
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus to the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{ScreenManager::display} in the main game loop calls
      \code{ScreenManager::update} when starting a new frame.
\item \code{ScreenManager::update} calls \code{screen->update} for the active
      game screen (i.e., a \code{GameSession} instance, considering the screen
      in which the \code{Bomber} appears).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{objs->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is indeed necessary considering the typical
OO game architecture employed in Pingus:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item With a single assignment to \code{screen}, one can easily deactivate the
current screen and redirect all dispatches to a new screen (step 2).
\item The \code{World} class manages and dispatches events to all game
      entities with a common interface \code{WorldObj}, such as the pingus and
      traps (step 4).
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus (step 5).
\item Since a single pingu can change its actions during lifetime, the
      \code{action} member decouples them with another level of indirection
      (step 6).
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions (step 7).
\end{itemize}
%
Like \code{update}, the \code{draw} callback also flows through a similar
dispatching hierarchy until reaching the \code{Sprite} class.

@@

In \CEU, the \code{Bomber} abstraction presented in Listing~\ref{lst.hier.b}
spawns a \code{Sprite} animation instance on its body (ln 3).
%
However, the \code{Sprite} abstraction in \CEU can react directly to external
\code{update} and \code{draw} events, bypassing the program hierarchy entirely.
Events in \CEU are broadcasted to the entire application in lexical order,
i.e., an abstraction that appears first in the source code (e.g., ln 3) reacts
before another one that appears second (e.g., hidden in ln 4).
%As discussed in Section~\ref{sec.pats.fsms.2}
This rule preserves determinism and also conforms to the program static
hierarchy.
While (\emph{and only while}) the bomber abstraction is alive, the sprite
animation remains alive and reacts to the \code{update} and \code{draw} events
(see also Section~\ref{sec.pats.lifespan} on Lifespan Hierarchies).
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.
% TODO: hidden in the language runtime
%Note that one can still declare a local event to restrict its visibility like
%a local variable.

@@

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
In addition, the \code{spawn} statement supports a pool identifier to associate
a new instance with a pool.
%
The game screen in Listing~\ref{lst.pingus.b} spawns a new \code{Pingu} on every
invocation of \code{Pingu\_Spawn} (ln 4--7).
%
The \code{spawn} statement (ln 6) specifies the pool declared at the top-level
block of the game screen (ln 3), attaching the scope of the pingu to that
block.
%In this case, the lifespan of the new instances follows the scope of the pool
%(ln 1--9) instead of the enclosing scope of the \code{spawn} statement
%(ln 4--7), surviving the call to \code{Pingu\_Spawn}.
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen (Figure~\ref{fig.pool}).
%
Lexical scopes handle memory and event dispatching automatically for static
instances (Listing~\ref{lst.hier.b}, ln 3) and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool.
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its \emph{natural termination}, the instance is automatically
removed from its pool ( Figure~\ref{fig.pool}, instances 1 and 3).
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (Listing~\ref{lst.pingus.b}, ln 17) aborts the
execution block of the \code{Pingu} instance, removing it from its associated
pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.
